---
title: 'Climate Change Module'
author: 'Quinn Thomas'
output:
  github_document: default
editor_options: 
  markdown: 
    wrap: 80
---

```{r message=FALSE}
library(tidyverse)
```

## Warm-up: Examining CO2 trends in R

- Example from <http://climate.nasa.gov/vital-signs/carbon-dioxide/>
- Raw data from <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt>

In 1958, Dr. Charles David Keeling (1928-2005), who was a scientist at Scripps Institute of Oceanography, began collecting data on atmospheric CO2 concentration at the Mauna Loa Observatory located in Hawaii. This dataset is what allowed us to understand the degree to which climate change is human-caused through our burning of fossil fuels and release of CO2 into the atmosphere. Due to his scientific achievements, Dr. Keeling was awarded the National Medal of Science by President George W. Bush in 2002. This is the highest award for lifetime scientific achievement that can be granted in the U.S. Today, you get to analyze this same dataset, except that you have more data that was available to Dr. Keeling and his colleagues, because your dataset extends up to current time.

```{r message=FALSE}

co2 <- 
read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date", "monthly_average",
                                "deseasonalized", "days", "sd_days","unc_month_mean"),
                  na = c("-1", "-99.99"))

```


```{r}
ggplot(co2, aes(x = decimal_date, y = monthly_average)) + 
  geom_line() + 
  geom_line(aes(y = deseasonalized), color = "blue") +
  labs(x = "Year", y = "CO2 concentration (ppm)") +
  theme_bw()
```

**Question 1:**

Describe the overall trend in the CO2 data.

**Answer 1: when plotting the monthly average CO2 concentrations (black solid line), the overall direction of CO2 concentrations is increasing; however, within each year, there are oscillations present, most likely representing seasons. Since seasons have widely different conditions (temperature, precipitation, light duration, etc), the oscillations seen are most likely due to these characteristics. When the data is smoothed out, by averaging the seasonal data for each year so the data is "deseasonalized", a clear linear increase can be seen (solid blue line) further indicating a positive direct relationship between year and CO2 concentrations. **

**Question 2:**

How does CO2 vary within a year?  What month is it at max?  Min?  What might explain this sub-annual pattern?

**Answer 2: In order to visualize the CO2 variations within a singular year, the CO2 concentrations of the years 2014 to 2022 were plotted and separated by year. All 9 plots show a very similiar trend/shape with the CO2 concentration increasing until around the first quarter of the year, where the concentration peaks, and then decreasing at relatively the same rate until the 3rd quarter of the year occurs, before finally increasing to a little more than the CO2 concentration as the beginning of the year. When examining the max and min CO2 concentrations of the whole data, May and June had the extreme maximum CO2 concentration while October had the extreme minimum. When examining the average of monthly averages of all years, May had the highest average with the surrounding months being second and third highest. **

```{r}
#CO2 variation within a year between 2014 and 2023
co2 |> filter(year >= 2014 & year < 2023) |> 
  ggplot(aes(x = decimal_date, y = monthly_average)) + 
  geom_line() +
  theme_bw() +
  labs(x = "Year", 
       y = "CO2 concentration (ppm)",
       title= "CO2 concentration (ppm) per Year from 2014 - 2022") +
  facet_wrap(~year, scales = "free") + 
  theme(axis.text.x = element_text(color = "grey20",
                                   angle=35, 
                                   vjust=0.5, 
                                   hjust=0.5))

#Month Max & Min (group_by)
co2 |>  group_by(month) |>  summarize(maxco2 = max(monthly_average, na.rm = TRUE))
co2 |>  group_by(month) |>  summarize(minco2 = min(monthly_average, na.rm = TRUE))
co2 |>  group_by(month) |>  summarize(meanco2 = mean(monthly_average, na.rm = TRUE))

```

## Global Temperature Data

Current climate change is affecting many aspects of the environment, with socio-economic consequences. For example, a warmer climate can allow new diseases to be introduced and persist (e.g. West Nile became established in the United States after an unusually warm winter allowed the mosquitoes that carry the virus to survive and spread). We are concerned not only with the actual temperature, but also with the rate that the temperature changes. Very rapid changes make it more likely that species cannot adapt and will go extinct. 

Each of the most recent years have the warmest on record.  In this section we will analyze global mean temperature data.

Data from: <http://climate.nasa.gov/vital-signs/global-temperature>

**Question 3:**

Describe the data set to the best of your ability given the documentation provided. Describe what kind of column each data contains and what units it is measured in. Then address our three key questions in understanding this data:

- How are the measurements made? What is the associated measurement uncertainty?
- What is the spatial resolution of the data? What is the temporal resolution of the data?
- Are there missing values? How should they be handled?

**Answer 3: This data set contains the change in global land-ocean temperature that occurred each year from 1880 to 2021, which were provided by NASA's Goddard Institute for Space Studies (GISS). There are two columns that contain change in temperature data with one being the actual average change in global temperature of the year and the other being the smoothed average change in global temperature of the year by 5 years. There is no stated uncertainty; however, NASA's analyses typically matched the independent studies by two other organizations: the Climatic Research Unit and the National Oceanic and Atmospheric Administration (NOAA). The spatial resolution came from satellite data, I assume since NASA was the organization that collected this data, and the temporal resolution was each year. It also does not appear that there are any missing values in this dataset. **


**Question 4:**

Construct the necessary R code to import and prepare for plotting the following data set: <http://climate.nasa.gov/system/internal_resources/details/original/647_Global_Temperature_Data_File.txt>

**Answer 4: See code block below. **

```{r}
#read_table
global_temp <- read_table("http://climate.nasa.gov/system/internal_resources/details/original/647_Global_Temperature_Data_File.txt", skip = 5, col_names = c("year", "gst_change", "lowess"))

```

**Question 5:**

Plot the trend in global mean temperatures over time.  Describe what you see in the plot and how you interpret the patterns you observe.

**Answer 5: When the plot is examined, it can be seen that average global surface temperature change is oscillating back and forth between small time frames, but typically increasing in overall trend. The oscillations appear to be constant around up until 1920 when the average change increases. The changes seem to level out until 1980 before rapidly increasing until 2021. I believe the oscillations may be due to the changing seasons or other factors. Either way, it is hard to analyze the overall trend, so the smoothed average for 5 years was plotted along the originial data. This way, the small decrease in the beginning and overall increase in the years following can be seen more clearly. **

```{r}
#just global surface temperature change
ggplot(global_temp, aes(year, gst_change)) + 
  geom_line() + 
  theme_bw() +
  labs(x = "Year", 
       y = "Change in Global Surface Temperature (C)", 
       title = "Global Mean Temperature Trend")

#global surface temperature change with smoothed average
global_temp |> ggplot(aes(year, gst_change)) + 
  geom_point() + 
  geom_line(aes(year, lowess)) + 
  theme_bw() +
  labs(x = "Year", 
       y = "Change in Global Surface Temperature (C)", 
       title = "Global Mean Temperature Trend")

```

## Evaluating the evidence for a "Pause" in warming?

The [2013 IPCC Report](https://www.ipcc.ch/pdf/assessment-report/ar5/wg1/WG1AR5_SummaryVolume_FINAL.pdf) included a tentative observation of a "much smaller increasing trend" in global mean temperatures since 1998 than was observed previously.  This led to much discussion in the media about the existence of a "Pause" or "Hiatus" in global warming rates, as well as much research looking into where the extra heat could have gone.  (Examples discussing this question include articles in [The Guardian](http://www.theguardian.com/environment/2015/jun/04/global-warming-hasnt-paused-study-finds), [BBC News](http://www.bbc.com/news/science-environment-28870988), and [Wikipedia](https://en.wikipedia.org/wiki/Global_warming_hiatus)). 

You will use rolling averages to help you explore the evidence for a pause. You have not been provided instructions for calculating rolling means so a learning objective of this question is to practicing finding the solution.

**Question 6:**

Use a search engine (e.g., Google), to search for how to calculate a rolling average in R. What search term did you use and what website did you end up using?  Why did you select this particular website?

**Answer 6: I searched "rolling averages in r" because I hoped to get a website that would first tell me what they are and their purpose, and then tell me how to calculate them in R. After searching this, I can across this site (https://www.statology.org/rolling-average-in-r/) titled "How to Calculate a Rolling Average in R (With Example)" which I chose because I could tell that it included the definition, how to calculate rolling averages, and had specific R codes I could use as a reference. **

**Question 7:**

- What is the meaning of "5 year average" vs "annual average"?
- Create a data frame from the annual data with three new columns: 5-year running averages, 10-year running averages, and 20-year running averages.

**Answer 7: Annual average refers to taking the mean of the data collected within one year. A 5-yr average refers to taking the mean of the data collected within 5 years, typically the current year and the 4 years before, but there are other ways to stagger the average as well. **

```{r}
#installing libraries
library(dplyr)
#install.packages("zoo")
library(zoo)

#adding rolling averages
## 5 years
global_temp <- global_temp |> mutate(rolling_ave5 = rollmean(gst_change, k = 5, fill = NA, align = "right"))
## 10 years
global_temp <- global_temp |> mutate(rolling_ave10 = rollmean(gst_change, k = 10, fill = NA, align = "right"))
## 20 years
global_temp <- global_temp |> mutate(rolling_ave20 = rollmean(gst_change, k = 20, fill = NA, align = "right"))

```

**Question 8:**

Plot the different averages on the same plot and describe what differences you see and why.  

**Answer 8: When the plot is examined, the three rolling average trends can be compared to the original data. As the rolling average time period increases, a clearer trend can be seen of the global surface temperature changes increasing. While "the pause" scientists were talking about could be seen on the original graph (as a seemingly flat portion of the graph), it can not be seen on the 20 year rolling average, and is less noticeable on the other rolling average graphs as well. This is because, when the data is averaged over a longer period of time, volitile data is less likely to affect the plot and trends can be viewed more accurately. **

```{r}
#plotting regular & rolling averages
global_temp |> ggplot(aes(year, gst_change)) + 
  geom_point() +
  geom_line(aes(year, rolling_ave5), color = "red") +
  geom_line(aes(year, rolling_ave10), color = "orange") +
  geom_line(aes(year, rolling_ave20), color = "blue") +
  theme_bw() +
  labs(x = "Year", 
       y = "Change in Global Surface Temperature (C)", 
       title = "Global Mean Temperature Trend")

```

**Question 9:**

By examining the data here, what evidence do you find or not find for such a pause? 

**Answer 9:**

## Longer term trends in CO2 Records

When analyzing Earth’s climate, it is important to remember that Earth is 4.54 billion years old. Our analyses so far have only looked at recent history. How can we compare the recent data to pre-historic time? Are the current rates of change similar or different than those the earth has experienced in the past? To explore this, we can use data taken from ice cores that were drilled at the poles. 

Hundreds of ice cores have been extracted from polar ice because they contain valuable data on atmospheric chemistry over pre-historic time. These valuable data exist in tiny air bubbles that are trapped in the ice. These air bubbles contain the same gases in the same ratios as the atmosphere at the time when the ice formed. The data you will be analyzing today are from ice cores extracted from the Vostok research station in Antarctica. As you have probably assumed, the depth of the ice core is related to how old the ice is; deep ice is older. There are two other variables that you will be analyzing from the ice cores.  You will analyze CO2 concentration, which has been measured from air bubbles trapped in the ice. We can use these data to see what rates of change were like during this pre-historic period, during which human activity has been minimal. 

[Ice core data](https://cdiac.ess-dive.lbl.gov/trends/co2/ice_core_co2.html):

Vostok Core, back to 400,000 yrs before present day 

- Description of data set: <https://cdiac.ess-dive.lbl.gov/trends/co2/vostok.html>
- Data source: <https://cdiac.ess-dive.lbl.gov/ftp/trends/co2/vostok.icecore.co2>

**Question 10:**

The broad question is: how do recent atmospheric CO2 levels compare to historical levels?

Your answers to this Question is going to be a mix of code chunks and text.

- Describe the data set: what are the columns and units? Where do the numbers come from? 
- What is resolution of the data? Interpretation of missing values?  
- Read in and prepare data for analysis.   
- Reverse the ordering to create a chronological record.    
- Plot data.  
- Consider various smoothing windowed averages of the data using in the plots.   
- Combine this time series with the Mauna Loa data (the Warm-up exercise).  Hint:  make sure both datasets have the same columns in the same order and use [`bind_rows`](https://dplyr.tidyverse.org/reference/bind_rows.html).   
- Plot the combined data.    
- Describe your conclusions to the question "how do recent atmospheric CO2 levels compare to historical levels?" using your plot as supporting evidence.   

**Answer 10:**

# Knitting and committing

Remember to Knit your document as a `github_document` and comment+push to GitHub your code, knitted document, and any files in the `figure-gfm` subdirectory that was created when you knitted the document.


